---
title: "bloodpressure"
output: html_document
---

Check the README for necessary R packages 

Load library's each session

```{r}
library(readxl)
library(ggplot2)
library(readr)
library(tidyr)
library(ez)
```


Read the data and assign to vector what you want to use. In this case we need the diastolic and systolic blood pressure of every person we measured. 

```{r}
data <- read_excel("../../raw_data/blood_pressure/bloodpressure.xlsx")

diastolic <- data$`Diastolic (mmHg)`
systolic <- data$`Systolic (mmHg)`
```


Check if there are any NA's in the data

```{r}
sum(is.na(diastolic))
sum(is.na(systolic))
```

Diastolic measurements

```{r}
rust_diastolic <- as.numeric(diastolic[seq(4, length(diastolic), by = 4)])

toets_diastolic_1 <- as.numeric(diastolic[seq(3, length(diastolic), by = 4)])
toets_diastolic_2 <- as.numeric(diastolic[seq(2, length(diastolic), by = 4)])
toets_diastolic_3 <- as.numeric(diastolic[seq(1, length(diastolic), by = 4)])
```


Systolic measurements


```{r}

rust_systolic <- as.numeric(systolic[seq(4, length(diastolic), by = 4)])

toets_systolic_1 <- as.numeric(systolic[seq(3, length(systolic), by = 4)])
toets_systolic_2 <- as.numeric(systolic[seq(2, length(systolic), by = 4)])
toets_systolic_3 <- as.numeric(systolic[seq(1, length(systolic), by = 4)])

```   


Means of systolic and diastolic measurements
```{r}
gem_rust_dia <- mean(rust_diastolic)
gem_toets_1_dia <- mean(toets_diastolic_1)
gem_toets_2_dia <- mean(toets_diastolic_2)
gem_toets_3_dia <- mean(toets_diastolic_3)


gem_rust_sys <- mean(rust_systolic)
gem_toets_1_sys <- mean(toets_systolic_1)
gem_toets_2_sys <- mean(toets_systolic_2)
gem_toets_3_sys <- mean(toets_systolic_3)
```


Gender and pss-10 score per contestant

```{r}
gender_pss_data <- read_csv("../../raw_data/google_forms/pss_gender_scores.csv")
gender <- gender_pss_data$Gender
pss_10 <- gender_pss_data$`PSS-10 (/40)`

```

Making 1 data frame for them all

```{r}
# Just choosing 1 of the measurements as all are measured once per contestant 
n <- length(rust_diastolic)
persoon <- 1:n

# Gender must be reversed as our first measurements of all the bloodpressure are our last measurements

bloodpressure_df <- data.frame(
  persoon = persoon,
  gender = rev(gender),
  pss_10 = rev(pss_10),
  
  rust_sys = rust_systolic,
  toets1_sys = toets_systolic_1,
  toets2_sys = toets_systolic_2,
  toets3_sys = toets_systolic_3,
  
  rust_dia = rust_diastolic,
  toets1_dia = toets_diastolic_1,
  toets2_dia = toets_diastolic_2,
  toets3_dia = toets_diastolic_3
)


bloodpressure_df
```


Line graph of the diastolic blood pressure during the rest phase and after the 3 tests

```{r}
df <- data.frame(
  meting  = c("Rust", "Toets 1 - (geen tijd)", "Toets 2 - (1 minuut)", "Toets 3 - (30 seconden)"),
  waarde  = c(gem_rust_dia, gem_toets_1_dia, gem_toets_2_dia, gem_toets_3_dia)
)

df$meting <- factor(df$meting, levels = c("Rust", "Toets 1 - (geen tijd)", "Toets 2 - (1 minuut)", "Toets 3 - (30 seconden)"))

ggplot(df, aes(x = meting, y = waarde, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Diastolische bloeddruk per meting",
       x = "Meting",
       y = "Diastolisch (mmHg)") +
  theme_minimal()


```


Line graph of the systolic blood pressure during the rest phase and after the 3 tests


```{r}

df <- data.frame(
  meting  = c("Rust", "Toets 1 - (geen tijd)", "Toets 2 - (1 minuut)", "Toets 3 - (30 seconden)"),
  waarde  = c(gem_rust_sys, gem_toets_1_sys, gem_toets_2_sys, gem_toets_3_sys)
)


df$meting <- factor(df$meting, levels = c("Rust", "Toets 1 - (geen tijd)", "Toets 2 - (1 minuut)", "Toets 3 - (30 seconden)"))

ggplot(df, aes(x = meting, y = waarde, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Systolische bloeddruk per meting",
       x = "Meting",
       y = "Systolisch (mmHg)") +
  theme_minimal()
```


Box graph of the diastolic blood pressure during the rest phase and after the 3 tests

```{r}

df_long_dia <- pivot_longer (
  bloodpressure_df,
  cols = c(rust_dia, toets1_dia, toets2_dia, toets3_dia),
  names_to = "meting",
  values_to = "waarde"
)

ggplot(df_long_dia, aes(x = meting, y = waarde)) +
 geom_boxplot() +
  labs(
       title = "Diastolische bloeddruk per meting",
       x = "Meting",
       y = "Diastolisch (mmHg)"
       ) 

```

Box graph of the diastolic blood pressure during the rest phase and after the 3 tests per gender

```{r}


ggplot(df_long_dia, aes(x = meting, y = waarde, fill = gender)) +
 geom_boxplot() +
  labs(
       title = "Diastolische bloeddruk per meting",
       x = "Meting",
       y = "Diastolisch (mmHg)",
       
       ) + 
       scale_fill_manual(values = c("Man" = "cyan", "Vrouw" = "lightcoral"))


```

Statistical verdict on our diastolic data

```{r}
shapiro.test(rust_diastolic)
qqnorm(rust_diastolic)

shapiro.test(toets_diastolic_1)
qqnorm(toets_diastolic_1)

shapiro.test(toets_diastolic_2)
qqnorm(toets_diastolic_2)

shapiro.test(toets_diastolic_3)
qqnorm(toets_diastolic_3)


# All data is normally spread, p > 0.05
# So repeated Anova test and then paired t test

# Sorts df_long on person so it prevents any mismatch
df_long_dia_sorted <- df_long_dia[order(df_long_dia$meting, df_long_dia$persoon), ]


result <- ezANOVA(
  data   = df_long_dia_sorted,
  dv     = waarde,
  wid    = persoon,
  within = meting
)

result$ANOVA

# Conclusion: 
# A repeated measures ANOVA showed no significant effect of measurement type on diastolic blood pressure,
# F(3, 57) = 2.11, p = .110, ges = .009, so we will not perform a paired T test

# Check for any significance between the genders
result_gender <- ezANOVA(
  data    = df_long_dia_sorted,
  dv      = waarde,
  wid     = persoon,
  within  = meting,
  between = gender
)

result_gender$ANOVA

# gender p = 0.055~ so not significant yet, but very borderline, worth mentioning in the final report that
# Woman tend to show a higher diastolic blood pressure than men.
# meting p = 0.105~ still the same, no significant difference across the measurements
# gender:meting p = 0.266~ no significant change, genders respond the same way across measurements. 

``` 



Box graph of the systolic blood pressure during the rest phase and after the 3 tests

```{r}

df_long_sys <- pivot_longer (
  bloodpressure_df,
  cols = c(rust_sys, toets1_sys, toets2_sys, toets3_sys),
  names_to = "meting",
  values_to = "waarde"
)

ggplot(df_long_sys, aes(x = meting, y = waarde)) +
 geom_boxplot() +
  labs(
       title = "Systolische bloeddruk per meting",
       x = "Meting",
       y = "Systolisch (mmHg)"
       ) 

```

Box graph of the systolic blood pressure during the rest phase and after the 3 tests per gender

```{r}

ggplot(df_long_sys, aes(x = meting, y = waarde, fill = gender)) +
 geom_boxplot() +
  labs(
       title = "Systolische bloeddruk per meting",
       x = "Meting",
       y = "Systolisch (mmHg)"
       ) +
      scale_fill_manual(values = c("Man" = "cyan", "Vrouw" = "lightcoral"))
        
```

Statistical verdict on our systolic data
```{r}

shapiro.test(rust_systolic)
qqnorm(rust_systolic)

shapiro.test(toets_systolic_1)
qqnorm(toets_systolic_1)

shapiro.test(toets_systolic_2)
qqnorm(toets_systolic_2)

shapiro.test(toets_systolic_3)
qqnorm(toets_systolic_3)

df_long_sys_sorted <- df_long_sys[order(df_long_sys$meting, df_long_sys$persoon), ]


result <- ezANOVA(
  data   = df_long_sys_sorted,
  dv     = waarde,
  wid    = persoon,
  within = meting
)

result$ANOVA

# P = 0.017~ < P = 0.05, so there's a significant statistical difference.
# Ges = 0.0142~ difference still small, but still significant.
# That's why we do a pairwise T test after this.

pairwise.t.test(
  df_long_sys_sorted$waarde,
  df_long_sys_sorted$meting,
  paired          = TRUE,
  p.adjust.method = "bonferroni"
)

# All p value's of 1.000 or above 0.05 are not significant. That there was a significant change
# Between rust_sys and toets2_sys and between toets2_sys and toets1_sys. Both P-value's < 0.05.
# Conclusion: toets 2 (1 minuut) caused a significant difference



t.test(toets_systolic_2, rust_systolic, paired = TRUE)
t.test(toets_systolic_2, toets_systolic_1, paired = TRUE)



# And for the gender of systolic bloodpressure
result_gender <- ezANOVA(
  data    = df_long_sys_sorted,
  dv      = waarde,
  wid     = persoon,
  within  = meting,
  between = gender
)

result_gender$ANOVA

# gender p = 0.116~ so not significantly different systolic bloodpressure overall between the 2 genders
# meting p = 0.020 still the same significant difference across the measurements (driven by toets2, see T test above)
# gender:meting p = 0.793 no significant change, genders respond the same way across measurements. 

```

We also want to see if people with a higher monthly experienced stress (measured with the pss-10)
Also experience an increase in blood pressure during the test, to try to measure this we will measure
the delta of test 1 to 3. By substracting their blood pressure during rest. 

```{r}
bloodpressure_df$delta_toets1_dia <- toets_diastolic_1 - rust_diastolic
bloodpressure_df$delta_toets2_dia <- toets_diastolic_2 - rust_diastolic
bloodpressure_df$delta_toets3_dia <- toets_diastolic_3 - rust_diastolic

bloodpressure_df$delta_toets1_sys <- toets_systolic_1 - rust_systolic
bloodpressure_df$delta_toets2_sys <- toets_systolic_2 - rust_systolic
bloodpressure_df$delta_toets3_sys <- toets_systolic_3 - rust_systolic

```


We will instantly check the statistical probability of the data used in the next 8 graphs
```{r}
shapiro.test(bloodpressure_df$pss_10)

shapiro.test(bloodpressure_df$delta_toets1_sys)
shapiro.test(bloodpressure_df$delta_toets2_sys)
shapiro.test(bloodpressure_df$delta_toets3_sys)

shapiro.test(bloodpressure_df$delta_toets1_dia)
shapiro.test(bloodpressure_df$delta_toets2_dia)
shapiro.test(bloodpressure_df$delta_toets3_dia)


```


The P value of every delta of every test is normal, except of test 2 for the diastolic blood pressure
That's why we run a spearman test for that instead of pearson.

```{r}
cor.test(bloodpressure_df$pss_10, bloodpressure_df$delta_toets1_sys, method = "pearson")
cor.test(bloodpressure_df$pss_10, bloodpressure_df$delta_toets2_sys, method = "pearson")
cor.test(bloodpressure_df$pss_10, bloodpressure_df$delta_toets3_sys, method = "pearson")

cor.test(bloodpressure_df$pss_10, bloodpressure_df$delta_toets1_dia, method = "pearson")
cor.test(bloodpressure_df$pss_10, bloodpressure_df$delta_toets2_dia, method = "spearman")
cor.test(bloodpressure_df$pss_10, bloodpressure_df$delta_toets3_dia, method = "pearson")
```




Diastolic blood pressure set out against the pss-10 scores (perceived stress scale, 10 questions) - test 1

```{r}
ggplot(bloodpressure_df, aes(x = pss_10, y = delta_toets1_dia)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "PSS-10 vs Diastolische bloeddruk (toets 1)",
    x     = "PSS-10 score",
    y     = "Bloeddruk (mmHg)"
  )

```

Diastolic blood pressure set out against the pss-10 scores (perceived stress scale, 10 questions) - test 2

```{r}
ggplot(bloodpressure_df, aes(x = pss_10, y = delta_toets2_dia)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "PSS-10 vs Diastolische bloeddruk (toets 2)",
    x     = "PSS-10 score",
    y     = "Bloeddruk (mmHg)"
  )
  
```

Diastolic blood pressure set out against the pss-10 scores (perceived stress scale, 10 questions) - test 3

```{r}
ggplot(bloodpressure_df, aes(x = pss_10, y = delta_toets3_dia)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "PSS-10 vs Diastolische bloeddruk (toets 3)",
    x     = "PSS-10 score",
    y     = "Bloeddruk (mmHg)"
  )
  
```


Systolic blood pressure set out against the pss-10 scores (perceived stress scale, 10 questions) - test 1

```{r}
ggplot(bloodpressure_df, aes(x = pss_10, y = delta_toets1_sys)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "PSS-10 vs Systolische bloeddruk (toets 1)",
    x     = "PSS-10 score",
    y     = "Bloeddruk (mmHg)"
  )
  
```


Systolic blood pressure set out against the pss-10 scores (perceived stress scale, 10 questions) - test 2

```{r}
ggplot(bloodpressure_df, aes(x = pss_10, y = delta_toets2_sys)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "PSS-10 vs Systolische bloeddruk (toets 2)",
    x     = "PSS-10 score",
    y     = "Bloeddruk (mmHg)"
  )
  
```


Systolic blood pressure set out against the pss-10 scores (perceived stress scale, 10 questions) - test 3

```{r}
ggplot(bloodpressure_df, aes(x = pss_10, y = delta_toets3_sys)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "PSS-10 vs Systolische bloeddruk (toets 3)",
    x     = "PSS-10 score",
    y     = "Bloeddruk (mmHg)"
  )
  
```
Added the graphs together as one as it has way more oversight.
And has a smaller footprint.

### Diastolic blood pressure

```{r}
df_delta_long_dia <- pivot_longer(
  bloodpressure_df,
  cols = c(delta_toets1_dia, delta_toets2_dia, delta_toets3_dia),
  names_to = "toets",
  values_to = "delta"
)

ggplot(df_delta_long_dia, aes(x = pss_10, y = delta)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~toets) +
  labs(
    title = "PSS-10 vs delta diastolische bloeddruk per toets",
    x     = "PSS-10 score",
    y     = "Stijging t.o.v. rust (mmHg)"
  )
  
```

### Systolic blood pressure

```{r}
df_delta_long_sys <- pivot_longer(
  bloodpressure_df,
  cols = c(delta_toets1_sys, delta_toets2_sys, delta_toets3_sys),
  names_to = "toets",
  values_to = "delta"
)

ggplot(df_delta_long_sys, aes(x = pss_10, y = delta)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~toets) +
  labs(
    title = "PSS-10 vs delta systolische bloeddruk per toets",
    x     = "PSS-10 score",
    y     = "Stijging t.o.v. rust (mmHg)"
  )
```



Now also if there's a difference between genders

Diastolic difference with the delta of each test and rest fase 

```{r}
df_delta_long_dia <- pivot_longer (
  bloodpressure_df,
  cols = c(delta_toets1_dia, delta_toets2_dia, delta_toets3_dia),
  names_to = "toets",
  values_to = "delta"
)


ggplot(df_delta_long_dia, aes(x = pss_10, y = delta, color = gender)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~toets) +
  labs(
    title = "PSS-10 vs delta diastolische bloeddruk per toets",
    x     = "PSS-10 score",
    y     = "Stijging t.o.v. rust (mmHg)",
    color = "Gender"
  ) + scale_color_manual(values = c("Man" = "cyan", "Vrouw" = "lightcoral"))

  
```


Systolic difference with the delta of each test and rest fase 

```{r}

df_delta_long_sys <- pivot_longer (
  bloodpressure_df,
  cols = c(delta_toets1_sys, delta_toets2_sys, delta_toets3_sys),
  names_to = "toets",
  values_to = "delta"
)

ggplot(df_delta_long_sys, aes(x = pss_10, y = delta, color = gender)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~toets) +
  labs(
    title = "PSS-10 vs delta systolische bloeddruk per toets",
    x     = "PSS-10 score",
    y     = "Stijging t.o.v. rust (mmHg)",
    color = "Gender"
    
  ) + scale_color_manual(values = c("Man" = "cyan", "Vrouw" = "lightcoral"))
  
  
```
 
Now the statistical analysis on the gender split diastolic and systolic blood pressure

First divide the data between the two genders in separate vectors
```{r}
mannen <- bloodpressure_df[bloodpressure_df$gender == "Man",]
vrouwen <- bloodpressure_df[bloodpressure_df$gender == "Vrouw",]
```



### Diastolic blood pressure:


Then check if the data is normally distributed and do this for all tests on diastolic bloodpressure
```{r}
shapiro.test(mannen$delta_toets1_dia)
shapiro.test(vrouwen$delta_toets1_dia)
shapiro.test(mannen$delta_toets2_dia)
shapiro.test(vrouwen$delta_toets2_dia)
shapiro.test(mannen$delta_toets3_dia)
shapiro.test(vrouwen$delta_toets3_dia)
```

All are normally distributed, but except for the women test1 and test2 so those get spearman for the correlation test
The rest are standard pearson.
```{r}
cor.test(mannen$pss_10, mannen$delta_toets1_dia)
cor.test(mannen$pss_10, mannen$delta_toets2_dia)
cor.test(mannen$pss_10, mannen$delta_toets3_dia)

cor.test(vrouwen$pss_10, vrouwen$delta_toets1_dia, method = "spearman")
cor.test(vrouwen$pss_10, vrouwen$delta_toets2_dia, method = "spearman")

cor.test(vrouwen$pss_10, vrouwen$delta_toets3_dia)
```
### Systolic blood pressure:

Then check if the data is normally distributed and do this for all tests on systolic bloodpressure

```{r}
shapiro.test(mannen$delta_toets1_sys)
shapiro.test(vrouwen$delta_toets1_sys)
shapiro.test(mannen$delta_toets2_sys)
shapiro.test(vrouwen$delta_toets2_sys)
shapiro.test(mannen$delta_toets3_sys)
shapiro.test(vrouwen$delta_toets3_sys)
```
All are normally distributed, but except for the women test1  so that gets spearman for the correlation test
The rest are standard pearson.
```{r}
cor.test(mannen$pss_10, mannen$delta_toets1_sys)
cor.test(mannen$pss_10, mannen$delta_toets2_sys)
cor.test(mannen$pss_10, mannen$delta_toets3_sys)

cor.test(vrouwen$pss_10, vrouwen$delta_toets1_sys, method = "spearman")

cor.test(vrouwen$pss_10, vrouwen$delta_toets2_sys)
cor.test(vrouwen$pss_10, vrouwen$delta_toets3_sys)
```

Nothing is significant enough all p > 0.05
